Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2022-01-25 concurrency/ITP 2022/Verification of Concurrent Binary Search Tree.tex
corrected some trivial typos

This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case we the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
\begin{align*}
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
&\texttt{acquire}(\ell');\\
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
&\texttt{release}(\ell);\\
&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{align*}
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

@@

VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications. Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic-spec}.  This is how we write pre- and postcondition in VST. 
\begin{figure}[htb]
\centering
\begin{verbatim}
Program Definition insert_spec :=
DECLARE _insert
ATOMIC TYPE W OBJ a INVS Ei Eo
WITH ...
PRE [ ... ]
  PROP (...)
  LOCAL (...)
  SEP (P_l) | (P_p)
POST [ ... ]
  PROP ()
  LOCAL ()
  SEP (Q_l) | (Q_p)
\end{verbatim}
\caption{Atomic Specification in VST}
\label{atomic-spec}
\end{figure}
\texttt{W} is the \textsf{TypeTree} representing the type of arguments passed with \texttt{WITH} clause; \texttt{a} is the abstract state for the triple; \texttt{Ei} and \texttt{Eo} are the sets of invariants names inside and outside the triple. The \texttt{PROP} clause describes things that that are true independent of program state, the \texttt{LOCAL} clause describes the values contained in C local variables, and the \texttt{SEP} clause represents the \emph{separating conjunction} (*) of \emph{spatial predicates}, predicates on some part of the memory. In VST, while proving that a function implements an atomic specification, the precondition will contains an $\mathsf{atomic\_shift}$ assertion with the public pre- and postcondition, and the masks inside it. This atomic shift can be accessed through following two rules:
$$\inference[\textsf{atomic\_commit}]{\forall a, R * P_p\ a \Rrightarrow \mathsf{EX}\ y,\ Q_p\ a\ y * R'\ y}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
$$\inference[\textsf{atomic\_rollback}]{\forall a, R * P_p\ a \Rrightarrow P_p\ a * R'}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R'}$$
During commit, we must provide resources $R$ which, in combination with the public precondition $P_p$, allow us to prove the public postcondition $Q_p$. Then, we gain access to an assertion $Q$ required by the postcondition of the function, and leftover resources $R'$. During rollback, we provide resources $R$ which, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as leftover resources $R'$. The rollback is specially used to learn some relationship between pieces of information (e.g. ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$. To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of roolbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.

>>>>

This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
\begin{align*}
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
&\texttt{acquire}(\ell');\\
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
&\texttt{release}(\ell);\\
&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{align*}
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

@@

VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications. Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic-spec}.  This is how we write pre- and postcondition in VST. 
\begin{figure}[htb]
\centering
\begin{verbatim}
Program Definition insert_spec :=
DECLARE _insert
ATOMIC TYPE W OBJ a INVS Ei Eo
WITH ...
PRE [ ... ]
  PROP (...)
  LOCAL (...)
  SEP (P_l) | (P_p)
POST [ ... ]
  PROP ()
  LOCAL ()
  SEP (Q_l) | (Q_p)
\end{verbatim}
\caption{Atomic Specification in VST}
\label{atomic-spec}
\end{figure}
\texttt{W} is the \textsf{TypeTree} representing the type of arguments passed with \texttt{WITH} clause; \texttt{a} is the abstract state for the triple; \texttt{Ei} and \texttt{Eo} are the sets of invariants names inside and outside the triple. The \texttt{PROP} clause describes things that that are true independent of program state, the \texttt{LOCAL} clause describes the values contained in C local variables, and the \texttt{SEP} clause represents the \emph{separating conjunction} (*) of \emph{spatial predicates}, predicates on some part of the memory. In VST, while proving that a function implements an atomic specification, the precondition will contains an $\mathsf{atomic\_shift}$ assertion with the public pre- and postcondition, and the masks inside it. This atomic shift can be accessed through following two rules:
$$\inference[\textsf{atomic\_commit}]{\forall a, R * P_p\ a \Rrightarrow \mathsf{EX}\ y,\ Q_p\ a\ y * R'\ y}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
$$\inference[\textsf{atomic\_rollback}]{\forall a, R * P_p\ a \Rrightarrow P_p\ a * R'}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R'}$$
During commit, we must provide resources $R$ which, in combination with the public precondition $P_p$, allow us to prove the public postcondition $Q_p$. Then, we gain access to an assertion $Q$ required by the postcondition of the function, and leftover resources $R'$. During rollback, we provide resources $R$ which, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as leftover resources $R'$. The rollback is specially used to learn some relationship between pieces of information (e.g. ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$. To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of rollbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.

### 2022-01-25 concurrency/ITP 2022/intro.tex
moved hand-over-hand to background, since it's already present in Gotsman et al.'s 2007 paper
corrected some trivial typos

@@

Our specifications use the logical atomicity introduced in the TaDA
logic \cite{tada}, in the form of \emph{atomic Hoare
triples}. Intuitively, an atomic Hoare triple $ \langle
P \rangle\,\mathbb{C}\, \langle P \rangle$ means the program
$\mathbb{C}$ ``atomically updates'' from $P$ to $Q$. The program may
actually take multiple steps, but evey step before the atomic update
(linearization point) must preserve the assertion $P$. Meanwhile, the
concurrent environment may also update the state before the
linearization point, as long as the states satisfies $P$. The
assertion $Q$ must become true at the linearization point, then the
environment can do whatever it likes. There is no guarantee that $Q$
is still preserved when $\mathbb{C}$ returns. For example, the
specification of our \texttt{insert} operation may be explained as
follows: during the execution of \texttt{insert}, there is
always \emph{some} BST, and at some point the \texttt{insert} will
take a BST $t$, insert a value with a certain key, and then eventually
returns (meanwhile other threads may have further modified the
inserted tree).

We employ the Verified Software Toolchain (VST) \cite{plfcc} to verify
the correctness of CBST. Although the concurrent separation logic
(CSL) has been formalized in VST rooted on the work of Hobor et
al.\ \cite{oraclesematic}, we extend it with two descendants of CSL so
as to accomplish the verification. One is the logical atomicity
mentioned above; the other is the higher-order ghost state in the
style of Iris \cite{higherorderghoststate}. The ghost states are used
to contruct both the global invariants and the local state in our
proofs. They will be further discussed in \S\ref{correctness}.

>>>>

% Contributions:
% verified C implementation of BST -- the first?
% technique for connecting locks with abstract state/atomicity
% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract
% lock-free implementation?

@@

Our specifications use the logical atomicity introduced in the TaDA
logic \cite{tada}, in the form of \emph{atomic Hoare
triples}. Intuitively, an atomic Hoare triple $ \langle
P \rangle\,\mathbb{C}\, \langle Q \rangle$ means the program
$\mathbb{C}$ ``atomically updates'' from $P$ to $Q$. The program may
actually take multiple steps, but every step before the atomic update
(linearization point) must preserve the assertion $P$. Meanwhile, the
concurrent environment may also update the state before the
linearization point, as long as the states satisfies $P$. The
assertion $Q$ must become true at the linearization point, then the
environment can do whatever it likes. There is no guarantee that $Q$
is still preserved when $\mathbb{C}$ returns. For example, the
specification of our \texttt{insert} operation may be explained as
follows: during the execution of \texttt{insert}, there is
always \emph{some} BST, and at some point the \texttt{insert} will
take a BST $t$, insert a value with a certain key, and then eventually
returns (meanwhile other threads may have further modified the
inserted tree).

We employ the Verified Software Toolchain (VST) \cite{plfcc} to verify
the correctness of CBST. Although the concurrent separation logic
(CSL) has been formalized in VST rooted on the work of Hobor et
al.\ \cite{oraclesematic}, we extend it with two descendants of CSL so
as to accomplish the verification. One is the logical atomicity
mentioned above; the other is the higher-order ghost state in the
style of Iris \cite{higherorderghoststate}. The ghost states are used
to construct both the global invariants and the local state in our
proofs. They will be further discussed in \S\ref{correctness}.
